Collaborative Visual Analysis on the Web with RCloud

### 2015-06-25 discussion.tex
some copy editing

Surprisingly, we found that when users realize that \emph{scratch work
is getting saved}, the extra versions become a mental burden for
them. Clearly, RCloud has now moved too far to the other side, and the
right solution is somewhere in the middle. An intriguing possibility,
suggested by one of our interviewees, is to implement auditing of data
analyses as it existed in S~\cite{Becker:1988:Auditing}, tying the
data artifacts to the processes that generated them. As a result, the
bookkeeping is reduced to selecting the \emph{data} to be kept, and
the code which generates it can be derived and saved transparently.

@@

\paragraph*{Discovery is hard} 
With hundreds of users and thousands of notebooks, the notebook tree
becomes unwieldy. The search function does not protect searchers from
obsolete and erroneous notebooks, which by default are retained
forever. In addition, some important metadata (has this notebook been
recently run, or edited?) is also not easily searchable. In other words,
\emph{textual search} is easy, but \emph{relevance search} is what we
need. Allowing users to \emph{tag} notebooks is a natural idea, but it
relies on users making a conscious effort, which goes somewhat against
our ideal of not asking the user to do work for the system. Rather, we
believe a better alternative is to have the system work for the user,
relying on passive metrics, such as how often a notebook has been run.
More generally, infrastructure for automatic \emph{curation} of the
software artifacts generated by teams seems to be an important avenue
for future work.

\paragraph*{Collaboration is hard}
While forking provides a simple way to continue someone else's work,
teams working together generate workflows for which forking is not
ideal. Although popular, forking poses a problem when dozens of
versions of a notebook proliferate. A fully general solution seems
intractable. Nevertheless, we expect future versions of the
notebook infrastructure to move toward git's ideal of decentralized
code repositories, instead of the exclusive ownership mode we
currently employ.

>>>>

Surprisingly, we found that when users realize that \emph{scratch work
is getting saved}, the extra versions become a mental burden for
them. Clearly, RCloud is an extreme, and the right solution is
in the middle. An intriguing possibility,
suggested by one of our interviewees, is to implement auditing of data
analyses as it existed in S~\cite{Becker:1988:Auditing}, tying
data artifacts to the processes that generated them. As a result,
bookkeeping can be reduced to selecting the \emph{data} to be kept,
and the code which generates it can be derived and saved transparently.

@@

\paragraph*{Discovery is hard} 
With hundreds of users and thousands of notebooks, a global notebook tree
becomes unwieldy. The search function does not protect searchers from
obsolete and erroneous notebooks, which by default are retained
forever. In addition, important metadata (has this notebook been
recently run, or edited?) is also not easily searchable. In other words,
\emph{textual search} is easy, but \emph{relevance search} is what we
need. Allowing users to \emph{tag} notebooks is natural, but it
relies on users making a conscious effort, which goes against
our ideal of not asking the user to do work for the system.
It is better to have the system work for the user,
relying on passive metrics, such as how often a notebook has been run.
More generally, infrastructure for automatic \emph{curation} of
software artifacts generated by teams seems to be an important avenue
for future work.

\paragraph*{Collaboration is hard}
While forking provides a simple way to continue someone else's work,
teams working together generate workflows for which forking is not
ideal. Although popular, forking poses a problem when dozens of
versions of a notebook proliferate. A fully general solution seems
intractable. Nevertheless, we expect future versions of the
notebook infrastructure to move toward git's ideal of decentralized
code repositories, instead of the exclusive ownership mode RCloud has now.

### 2015-06-25 introduction.tex
some copy editing

RCloud knits together some familiar tools, and 
provides new features to find data and code; create experimental workbooks;
run experiments; annotate and deploy experiments as end-user websites or
as reusable, callable services; and to share, search and recommend these artifacts. The artifacts are stored in a version
control system that provides a common workspace, as well as needed control
and isolation between stable and experimental versions of code and other
resources.

A key point is that, for the most part, the improved capabilities are
available by default, without much explicit involvement on the part of
data scientists and other customers. Visual analytics experiments are
conveniently shared and turned into production websites, without moving
or porting code. All the published artifacts in the system can be
searched immediately. Recommendation is as easy as clicking a star
on a useful workbook.

The high level architecture of the prototype system
is shown in figure~\ref{fig:system}.
We chose R as the foundation for our prototype because it is already
the dominant statistical computing language in our lab.
R also has many useful packages for data analysis and visualization,
and the core system and its packages are open source that can be modified to
support research.

>>>>

RCloud knits together some familiar tools, and 
provides new features to find data and code; create experimental notebooks;
run experiments; annotate and deploy experiments as end-user websites or
as reusable, callable services; and to share, search and recommend these artifacts. The artifacts are stored in a version
control system that provides a common workspace, as well as needed control
and isolation between stable and experimental versions of code and other
resources.

A key point is that, for the most part, the improved capabilities are
available by default, without much explicit involvement on the part of
data scientists and other customers. Visual analytics experiments are
conveniently shared and turned into production websites, without moving
or porting code. All the published artifacts in the system can be
searched immediately. Recommendation is as easy as clicking a star
on a useful notebook.

The high level architecture of the prototype system
is shown in figure~\ref{fig:system}.
We chose R as the foundation because it is already
the dominant statistical computing language in our lab.
R also has many useful packages for data analysis and visualization,
and the core system and its packages are open source that can be modified to
support research.

### 2015-06-25 relatedwork.tex
figure squeezing
some copy editing

\begin{figure*}
\centering
\includegraphics[width=.75\linewidth]{fig/system/system.pdf}
\caption{\label{fig:system}A diagram of RCloud's architecture. }
\end{figure*}

@@

\paragraph*{Social Data exploration and Analysis}
ManyEyes~\cite{Viegas:2007:MAS} was a landmark system for the
crowdsourced creation and publication of data visualizations.
Although ManyEyes supported only a fixed set of visual encodings,
the system's success was an early indication of the potential to
combine social media on the web with visual analytics.

\paragraph*{Notebooks as a medium for data analysis dissemination}
The concept of a ``notebook'' as we use it can be traced
back to Knuth's literate programming~\cite{Knuth:1984:LP}. In
literate programming, a prose description of the behavior of
a program is ``weaved'' with its source code, yielding both an
executable program and a human-readable document.
A notebook represented as a collection of small executable cells,
originated with Mathematica, and in R, literate programming is
supported by packages such as knitr and RMarkdown~\cite{Xie:2013:DDW}.
Project Jupyter~\cite{jupyter} (originally implemented as part of
IPython) offers some notebook features, though it lacks a transparent
mechanism for multi-user sharing and deployment.
%
Further afield, {Electronic Lab Notebooks} are applications for organizing
and sharing data from scientific lab experiments\cite{Rubacha:2011:ELN}.
In a sense, we hope to adapt and extend this concept to the work of
visual analytics teams.
%
\stephen{RStudio is highly acclaimed for its...}
Although RStudio offers publication of literate R programs as a free
service on their website, the workflow is somewhat disconnected from
the development of those programs. Once they're uploaded, it's hard
for other users to build off of the work published (or even for the
original author to update new versions). In other words, RStudio
handles \emph{publication}, but not \emph{deployment} and sharing.

\paragraph*{Provenance and versioning} As
mentioned, one central issue in exploratory analysis is that
problems can change quickly, often in the course of developing a
solution. As a result, systems should provide adequate support for
tracking \emph{changes} of the data analysis scripts. VisTrails was
one of the original systems for managing \emph{process provenance} ,
and demonstrated the value of capturing aspects of the processes that
surround data analysis experiments and tools, including detailed
history, collaboration, and deployment~\cite{Callahan:2006:VVM}.

@@

\paragraph*{Web-based tools for sharing code snippets}
There have been quite a few tools recently developed for quickly
sharing small programs on the web, including
bl.ocks~\cite{blocks}, jsfiddle~\cite{jsfiddle}, and
plot.ly~\cite{plotly}. bl.ocks and jsfiddle are designed to
share Javascript programs, which means that deployment happens
automatically through the web browser. If Javascript eventually becomes
the lingua franca of exploratory data analysis, we can foresee
building a simpler version of RCloud where all execution happens
either on the client side or via web services. Unfortunately that is
not a realistic assumption at present, making these stand-alone
solutions unsuitable for our purpose. Plot.ly is notable in that it
provides API support for publishing \emph{from} scripts: in other
words, it is possible to generate a plot.ly visualization from inside
another program. Although this is an intriguing idea, it
nevertheless creates a disconnection between the analysis and the
resulting visualization. In RCloud, we wanted to ensure that every
visualization is transparently linked to the source code that
generates it.

\paragraph*{Needs of data analysts}
Kandel et al.'s interview study points out the typical ``explore'',
``model'', ``report'' cycle in enterprise data
analysis~\cite{Kandel:2012:EDA}. There are many discontinuities in
this cycle that cost time and effort to overcome. RCloud seeks to
reduce this mismatch. Kandel et al also point out that larger teams
are becoming more common in data analysis, that supporting
collaboration is difficult and important, and that sharing
and versioning of data sources and artifacts is hindered by current
technology. ``We found that analysts typically did not
share scripts with each other. Scripts that were shared were
disseminated similarly to intermediate data: either through shared
drives or email. Analysts rarely stored their analytic code in source
control.'' Their study highlights the opportunity for better ways
of supporting collaboration and sharing in data analysis teams.

@@

One overall goal is to reduce the gap between implementers and
deployers in visual analytics. The fusion of development with
production operations in software release management (``DevOps''
\cite{Httermann:2012:DD} or ``continuous integration''
\cite{Fowler:2006:Continuous}) is a trend in web services and similar
fields. By making it convenient for data scientists to expend just a
little additional effort when creating experiments, we may be able to
eliminate the need for programming teams to recreate their work to
deploy it in production, which has a high cost in time, expense and
accuracy.

We next give a description of the system architecture,
and how it enables capabilities that satisfy the requirements as described.

>>>>

\begin{figure*}
\centering
\includegraphics[width=.9\linewidth]{fig/system/system.pdf}
\caption{\label{fig:system}A diagram of RCloud's architecture. }
\end{figure*}

@@

\paragraph*{Social Data exploration and Analysis}
ManyEyes~\cite{Viegas:2007:MAS} was a landmark system for the
crowdsourced creation and publication of data visualizations.
Although ManyEyes supported only a fixed set of visual encodings,
the system's success was an early indication of the potential to
combine social media with visual analytics.

\paragraph*{Notebooks as a medium for data analysis dissemination}
The concept of a ``notebook'' as we use it can be traced
back to Knuth's literate programming~\cite{Knuth:1984:LP}.
In literate programming, a prose description of the behavior of
a program is ``weaved'' with its source code, yielding both an
executable program and a human-readable document.
A notebook represented as a collection of small executable cells,
originated with Mathematica, and in R, literate programming is
supported by packages such as knitr and RMarkdown~\cite{Xie:2013:DDW}.
Project Jupyter~\cite{jupyter} (originally implemented as part of
IPython) offers some notebook features, though it lacks a transparent
mechanism for multi-user sharing and deployment.
%
Further afield, {Electronic Lab Notebooks} are applications for organizing
and sharing data from scientific lab experiments\cite{Rubacha:2011:ELN}.
In a sense, we hope to adapt and extend this concept to the work of
visual analytics teams.
%
RStudio is highly acclaimed for its polished and powerful human interface.
Although RStudio offers publication of literate R programs as a free
service on its website, the workflow is somewhat disconnected from
the development of those programs. Once they're uploaded, it's hard
for other users to build off the work published, or even for the
original author to update new versions. In other words, RStudio
handles \emph{publication}, but not \emph{deployment} and sharing.

\paragraph*{Provenance and versioning} As
mentioned, one central issue in exploratory analysis is that
problems may change quickly, often while in the course of
developing a solution. As a result, systems need to provide adequate
support for tracking \emph{changes} of data analysis scripts. VisTrails
was one of the original systems for managing \emph{process provenance},
and it demonstrated the value of capturing aspects of the processes that
surround data analysis experiments and tools, including detailed
history, collaboration, and deployment~\cite{Callahan:2006:VVM}.

@@

\paragraph*{Web-based tools for sharing code snippets}
Quite a few tools were recently developed for quickly
sharing small programs on the web, including
bl.ocks~\cite{blocks}, jsfiddle~\cite{jsfiddle}, and
plot.ly~\cite{plotly}. bl.ocks and jsfiddle are designed to
share Javascript programs, which means that deployment happens
automatically through the web browser. If Javascript eventually becomes
the lingua franca of exploratory data analysis, we foresee
building a simpler version of RCloud where all execution happens
either on the client side or via web services. Unfortunately that is
not a realistic assumption at present, making these stand-alone
solutions unsuitable for our purpose. Plot.ly is notable in that it
provides API support for publishing \emph{from} scripts: in other
words, it is possible to generate a plot.ly visualization from inside
another program. Although this is an intriguing idea, it
nevertheless creates a disconnection between the analysis and the
resulting visualization. In RCloud, we wanted to ensure that every
visualization is transparently linked to the source code that
generates it.

\paragraph*{Needs of data analysts}
Kandel et al.'s interview study points out the typical ``explore'',
``model'', ``report'' cycle in enterprise data
analysis~\cite{Kandel:2012:EDA}. There are many discontinuities in
this cycle that cost time and effort to overcome. RCloud seeks to
reduce this mismatch. Kandel et al also point out that larger teams
are becoming more common in data analysis, that supporting
collaboration is difficult and important, and that sharing
and versioning of data sources and artifacts is hindered by current
technology. ``We found that analysts typically did not
share scripts with each other. Scripts that were shared were
disseminated similarly to intermediate data: either through shared
drives or email. Analysts rarely stored their analytic code in source
control.'' Their study highlights the opportunity to contribute better
ways of supporting collaboration and sharing by data analysis teams.

@@

One high-level goal is to reduce the gap between implementers and
deployers in visual analytics. The fusion of development with
production operations in software release management (``DevOps''
\cite{Httermann:2012:DD} or ``continuous integration''
\cite{Fowler:2006:Continuous}) is a trend in web services and similar
fields. By making it convenient for data scientists to expend just a
little additional effort when creating experiments, we may be able to
eliminate the need for programming teams to recreate their work to
deploy it in production, which has a high cost in time, expense and
accuracy.

% We next give a description of the system architecture,
% and how it enables capabilities that satisfy the requirements as described.
% \stephen{This should flow naturally, or have a more natural transition.}
